Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
unlock 6 7

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as \any.

@@

%
%
%
\section{Left overs}

SOME OF THIS STUFF CAN BE USED IN THE EVALUATION

%
%
\subsubsection{Possibly-NA Primitives}

A common primitive usage pattern is related to the presence of NAs:
programmers explicitly check for NAs in primitive vectors, possibly
sanitizing them if they are present.  NAs complicate practically every
computation using primitive vectors, a fact that many programmers appear to
be acutely aware of. For instance, consider the code in
Figure~\ref{fig:na-example}.

\begin{figure}[htbp]
\begin{center}

@@

\caption{NA checking example from the \code{binom} package.}
\label{fig:na-example}
\end{center}
\end{figure}

The \code{binom} package is a popular R package for computing confidence
intervals for binomial experiments, and the \code{binom.profile} function
does so using the profile likelihood, the details of which are not relevant
here.  This code snippet highlights a data sanitization pattern appearing at
the beginning of the function: the programmer first binds the vectors into a
matrix (line 2), with one column for each of \code{x} and \code{n}, then
finds rows where both columns are not NA (line 3), then extracts non-NA
values and stores them into \code{x} and \code{n} respectively (lines 4-5).
Importantly, this process is entirely quiet, and the user of
\code{binom.profile} will not be notified if their data happens to contain
NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
in data could signify some sort of data corruption, and quietly ignoring
issues in data could lead users (read: statisticians) to draw flawed
conclusions from erroneous data.

To combat this, we introduce a modifier on primitive types: the type
\code{^T} signifies that primitive type \code{T} might contain an NA value,
while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
shorter syntax as NA-free vectors are \AT{more common} than the alternative.
If the \code{binom} developers were to annotate arguments \code{x} and
\code{n} with such an NA-free type, \contractr will notify users when
\code{binom.profile} is called with NA-full values, rather than just quietly
ignoring them.

Note that when combining call traces involving \code{^T} and \code{T} types,
we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
vectors are passed to an argument, then we say that the argument can handle
NA-full vectors.  This is equivalent to saying \code{T <: ^T}.

%
%
\subsubsection{Scalar Primitives}

Initial analyses of our data revealed that programmers \AT{often} use
scalars (as explicitly as they can, anyway), and often do high-level
dimensionality checks on their data.  To illustrate, consider the code in
Figure~\ref{fig:scalar-vector-example}.

\begin{figure}[htbp]
\begin{center}

@@

\caption{Scalar/vector example from the \code{matrixcalc} package.}
\label{fig:scalar-vector-example}
\end{center}
\end{figure}

The \code{hankel.matrix} function takes two arguments (as described in the
comment in the code), and returns a Hankel matrix, which essentially is a
matrix where the skew-diagonals have a desirable property.  The function
body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
vector, the code \code{m < n} would essentially check \code{m < n[i]} for
each element in \code{n}, returning a vector of comparison results, and the
conditional of an if must be a scalar else R generates a warning.

These checks exemplify functionality that we would like to include in our
type annotation language, namely, we would like our type language to
differentiate between scalars and vectors.  To that end, we introduce the
following types into our type language: for some primitive type \code{T},
the user may modify \code{T} with a modifier \code{[]} to indicate that it
is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
\AT{occurred more frequently} than vectors, and for that reason we chose the
shorter scalar type syntax as the default.  The NA-free modifier works as
expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
while \code{^T} indicates a scalar which is not NA.

%
%
\subsubsection{Matrices}

Matrices represent another layer of complexity on top of vectorized
primitives.  Internally, matrices are simply vectors with class ``matrix''
and a \code{dims} attribute indicating matrix dimensions, and while not
codified in the language semantics, many internal functions will coerce
vectors to matrices automatically.  Relying on automatic coercion is unwise,
and many functions check to ensure that their arguments are matrices.  For
instance, consider the code in Figure~\ref{fig:matrix-example}.

\begin{figure}[htbp]
\begin{center}

@@

\caption{Matrix example from the \code{matrixStats} package.}
\label{fig:matrix-example}
\end{center}
\end{figure}

@@

%
%
\subsubsection{Forgoing Dimensionality}

Thus far we've seen instances where we introduce types that subsume existing
argument validation code, but our intention is not to subsume {\it all} such
code.  As an example of something we {\it do not} want to do, consider the
code in Figure~\ref{fig:general-validation-example}.

\begin{figure}[htbp]
\begin{center}

@@

\caption{General dimensionality check example from the \code{matrixStats} package.}
\label{fig:general-validation-example}
\end{center}
\end{figure}

@@

%
%
%
%
%
\subsubsection{Relationship Between Lists and Vectors}

The ease with which vectors may be transformed into lists is perhaps
indicative of a relationship between the two.  In R, all vectors can be
converted to lists with a simple call to \code{as.list}, and lists to
vectors with \code{unlist}, though care must be taken when unlisting a
non-homogeneous list, as coercions will be performed where possible, and the
function will simply do nothing if coercion is impossible.  That said, lists
and vectors have different access syntax, so vectors cannot stand-in for
lists without the \code{as.list} call above.

In sum, we opted not to include a subtyping relationship between vectors and
lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
    | list<T>} types occur?}.  During our evaluation we also found it rare
that programmers would attempt to pass list-typed values to vector-typed
arguments and vice versa, and we will discuss this further in
Section~\ref{sec:evaluation}.

%
%
%
\subsubsection{Structs}
\label{subsec:structs}

We observed the use of consistently-named lists in \AT{a number of
  functions}.  In R, ``fields'' of named lists can be accessed using the
\code{$} operator, as illustrated in the code snippet in
Figure~\ref{fig:struct-ex}.

\begin{figure}[htbp]
\begin{center}

@@

\caption{Struct example from a test in the \code{agricolae} package.}
\label{fig:struct-ex}
\end{center}
\end{figure}

@@

We still believe struct types to be useful to R programmers, and so our
language supports the type and \contractr will check them, though we
configured \typetracer to ignore them as they made the data substantially
less useful.  On the positive side, we found that \AT{get the number, but
  often} structs-typed values had a {\it class}, which helped motivate our
inclusion of class-like types in our type annotation language.  Classes are
discussed in more detail in Section~\ref{subsec:classes}.

%
\subsubsection{Data Frames}

One of the most popular classes in R is the \code{data.frame} class \AT{we
  have numbers here, grab em}.  Data frames and the derivative \code{tibble}
(from the \code{tidyverse} ecosystem) and \code{data.table} (from the
\code{data.table} package) types underpin a huge amount of data analysis in
R.

One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise into \typetracer's analysis results.  Further
complicating data frames is that many functions built to operate on them
operate in a name-agnostic way.  For instance, the \code{tidyverse} package
ecosystem allows programmers to pass column names to functions which operate
on their data frames.  In base R, idiomatic data frame use is to use string
column names to select rows from the frame (unless only a single column is
of interest, wherein the \code{$} syntax is appropriate).  We include some
examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.

\begin{figure}[htbp]
\begin{center}

@@

\caption{Select examples of idiomatic \code{data.frame} use.}
\label{fig:data-frames-bad}
\end{center}
\end{figure}

@@

%
%
%
%
\subsection{Classes in R}
\label{subsec:classes}

As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
of type.  The classes of function arguments are used by R as part of its
dispatch mechanism, the most prevalent of which (\AT{we have that number,
  too!}) is S3 dispatch, in which R dispatches on the (S3) class of the
first argument to a function, selecting the appropriate method to call for
the class.  To quickly illustrate, consider the code snippet in
Figure~\ref{fig:dispatch-ex}.

\begin{figure}[htbp]
\begin{center}

\begin{lstlisting}
# Quick example of S3 dispatch.
\end{lstlisting}

\caption{S3 dispatch in R.}
\label{fig:dispatch-ex}
\end{center}
\end{figure}

Classes are very prevalent in R \AT{number}, and so we include a
\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
class of values, but some care needed to be taken so as to not overreport
class types, as {\it all} values have a class in R.  The main classes we
were interested in were matrix, array (higher dimensional matrices), and
data.frames, in addition to any user-defined classes.  In many cases
\AT{number?}, values with a user-defined class are in reality lists with
named elements, or structs.  Including class types, and having \typetracer
report them in place of structs, significantly reduced the noise in the
analysis and allowed us to gather more focused data.  To get a sense for
this difference, consider the code in Figure~\ref{fig:class-vs-struct}.

\begin{figure}[htbp]
\begin{center}

\begin{lstlisting}
# Example of gross struct that turns into class.
\end{lstlisting}

\caption{S3 dispatch in R.}
\label{fig:class-vs-struct}
\end{center}
\end{figure}

When set to ignore classes, \typetracer reports \AT{type} as the type of the function, as compared with \AT{other type} when configured to report classes.

%
%
\subsubsection{Multi-Classes}

A value can have multiple classes in R.  This is common practice in the
\code{dplyr} package, which extends the functionality of data frames: a
\code{data.frame} that passes through \code{dplyr} functions gains two
additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
least 3, in addition to whichever other classes that data frame happened to
have.  Note that when looking to perform S3 dispatch, R will try to match
{\it any} of the classes of the first argument to a dispatched function, so
each of these classes are important.

We found these so-called multi-classes to be surprisingly common, making up
\AT{occurrence} of the occurrences of class-typed values.  One possible
reason for this is that values can gain classes quite easily (e.g. the
\code{dplyr} example discussed above), somewhat polluting the output of
\typetracer.  Nevertheless, we found \AT{number} of occurrences of
monomorphic multi-class arguments, meaning that the function was always
called with a multi-class value in that argument position.

In the end, we support these multi-class types by allowing \code{class<...>}
types to contain a list of class names, so for instance a \code{dplyr}
\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
type language, and when performing related contract assertions \contractr
will check that the value has every class that make up its type.

%
%
\subsubsection{Forgoing R's Object Systems}

@@

%
%
\section{Evaluation}
\label{sec:evaluation}

To evaluate the effectiveness of our type annotation language and contract
checking framework, we conducted a large-scale experiment based on the
corpus of 412 packages we used to inform our language design.  This corpus
represents some of the most widely used and thoroughly tested R libraries on
CRAN.

We configured \typetracer to infer the types of each of the functions that
are visible to package clients.  Recall that \typetracer's primary function
is to collect call signatures from observed function calls, and we ran
\typetracer on the test, example, and vignette code of all packages in the
aforementioned corpus.  These call traces were consolidated into a single
function type for each visible package function: \typetracer employs
unification strategies discussed in Section~\ref{sec:typesystemdesign} to
keep the size of signatures in check and only generates an \code{any} type
when missing and unused arguments were recored, and when a longjump caused
function execution to abort prematurely.  The types generated by \typetracer
will be analyzed in more depth in Section~\ref{subsec:stats}, and a
large-scale evaluation of the quality of these types will be discussed in
Section~\ref{subsec:eval}.

%
%
%
%
\subsection{Language Statistics}
\label{subsec:stats}

\subsection{Is our simple type language expressive enough to type R code?} Functions in
R can support arguments of multiple types by reflecting on the type to
perform appropriate conversions and computations. On the one hand, this form
of ad-hoc polymorphism makes R functions expressive; on the other hand, it
requires support for polymorphism in the typing scheme, making the type
system complex.  Our system lacks explicit support for
polymorphism. Instead, it supports union, \any and \VARGS types. A union type is
inferred when values of multiple types are observed at a function parameter
and return position, \any type is inferred when an argument is missing or
unused, and \VARGS is inferred for variable length argument list positions. Furthermore, list element
types can be inferred as \any if they have more than 5 different types. Values with many different types can result in
a large number of alternatives in the type inferred for that position,
making it hard to read and unwieldy for end-users. To evaluate this
''expressivity gap`' between our type inference scheme and R functions, we
look at the number of different types in the parameter and return types of
the XXX functions in our corpus.

\begin{figure}[!h]  \centering
  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
  \caption{Types inferred for each position}\label{fig:unionfreq}
\end{figure}

Figure XXX shows ten most common type categories that occur by themselves. We
observe that three scalar types, double, logical, and character are most
frequent. This is followed by XXX vararg parameter positions. Next, XX% of the
parameter types are any because arguments at those positions were either missing
or unused in at least one call to those functions. Together, these types amount
to more than 50%. This is followed by double vector, the unit type, null,
character vector, and classes matrix and data.frame respectively.

class is the most common type in our corpus, accounting for XXX% of all the
inferred types. This is followed by an almost equal proportion,
28%, scalar types and, surprisingly, only 13% vector types. Vectors are the
workhorse of R, most primitive operations in R are vectorized and literals are
treated as single element vectors. It is unexpected to observe twice as many
scalars as vectors. It is even more unexpected to observe class as the most
common type. class of an object is used for dynamic dispatch by the S3 object
system in R. This suggests that usage of S3 is prevalent in our corpus.

null types. Together, these four types make up for 80% of the observed types in
our corpus.
The other types, occurring only XXX% of the times include XXX environment types,
XXXX expression types, XXXX pairlist types and lastly, XXX externalptr types.

We analyzed the type signatures generated by \typetracer in order to build
up a profile of our type annotation language.  We were interested in broad,
high-level information, such as the distribution of generated types, the
level of polymorphism observed in the types, and the distribution of sizes
of union types.

%
%
\subsubsection{Type Distribution}

\AT{Include a figure? Histogram(s) of types?}

In order to count the occurrences of each type, we subdivided them into {\it kinds}: 
For our purposes, the {\it kind} of a type is equivalent to it's top-level type constructor, so all classes have kind ``class'', all lists have kind ``list'', and so on.
We found that classes are the most common kind of type, accounting for roughly 31\% of types.
\AT{Make sure to transform class<function> into any=>any.}
In order, the most common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%), factors (2\%), and tibbles (2\%).
Roughly 25\% of classes are part of R's base language implementation, the others being defined as part of some package's functionality.

It would appear that our simple type annotations nicely account for the landscape of values in R.
\AT{Not sure what to say about classes.}
% Classes are a common kind of type in R, and for the purposes of this type annotation language allowing users to specify class names as part of class types is welcome: we saw that the breakdown of the class kind does not reveal many outlier classes,

%
%
\subsubsection{Polymorphism}

\AT{We need better numbers on this.}
We found that 15\% of argument positions were polymorphic, and that X\% of functions were polymorphic.

%
%
\subsubsection{Size of Union Types}

%
%
\subsubsection{Use of \code{any}}
\label{subsec:any}

%
%
%
%
\subsection{RQ2: How good are type signatures generated from traces?}
\label{subsec:eval}

In the last section, we mainly discussed the function types that \typetracer generated, focusing on analyzing the polymorphism present in the types.
Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies.
We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.

%
%
\subsubsection{Overall Results}

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF of assertions failing: this suggests that the simplification we made did not greatly affect the results.
Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
These S3 functions represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

%
%
\subsubsection{Breakdown of Contract Assertion Failures}

We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with \VEC\dbl type has type \CLASS{matrix}.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a \VEC\dbl has type \VEC\int, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed, and overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework: thus, these functions implement user-defined polymorphism which we do not handle in our type framework.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.

>>>>

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as
\LIST{\any}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\FIX}[1]{{\bf #1}\xspace}

\subsection{Expressiveness}

The first part of our evaluation attempts to shed light on how good a fit
our proposed type system is with respect to common programming patterns
occuring in widely used R libraries.

First we look at the share of monomorphic arguments and function signatures.
Monomorphic in this context means that the type is not relying on \any or
inlcuding a union.  The import of monorphism in this context is that it
means our type language can accurately capture an argument's type or a
function signature.

<<Insert proportion of monorphic types in argument and in function
signature. Perhaps a table of monomorphic type signatures that are
popular...>>

\begin{figure}[!h]  \centering
  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
  \caption{Size of unions}\label{fig:unionfreq}
\end{figure}

Figure XXX shows ten most common type categories that occur by
themselves. We observe that three scalar types, double, logical, and
character are most frequent. This is followed by XXX vararg parameter
positions. Next, XX\% of the parameter types are any because arguments at
those positions were either missing or unused in at least one call to those
functions. Together, these types amount to more than 50\%. This is followed
by double vector, the unit type, null, character vector, and classes matrix
and data.frame respectively.

class is the most common type in our corpus, accounting for XXX\% of all the
inferred types. This is followed by an almost equal proportion,
28\%, scalar types and, surprisingly, only 13\% vector types. Vectors are the
workhorse of R, most primitive operations in R are vectorized and literals are
treated as single element vectors. It is unexpected to observe twice as many
scalars as vectors. It is even more unexpected to observe class as the most
common type. class of an object is used for dynamic dispatch by the S3 object
system in R. This suggests that usage of S3 is prevalent in our corpus.

null types. Together, these four types make up for 80\% of the observed
types in our corpus.  The other types, occurring only XXX\% of the times
include XXX environment types, XXXX expression types, XXXX pairlist types
and lastly, XXX externalptr types.

In order to count the occurrences of each type, we subdivided them into {\it
  kinds}: For our purposes, the {\it kind} of a type is equivalent to it's
top-level type constructor, so all classes have kind ``class'', all lists
have kind ``list'', and so on.  We found that classes are the most common
kind of type, accounting for roughly 31\% of types.  In order, the most
common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%),
factors (2\%), and tibbles (2\%).  Roughly 25\% of classes are part of R's
base language implementation, the others being defined as part of some
package's functionality.

We found that 15\% of argument positions were polymorphic, and that X\% of
functions were polymorphic.

\subsubsection{Discussion}

\paragraph{NAs}
Our data supports the choice of making the presence of NAs explicit.  In
practice, programmers explicitly check for NAs in primitive vectors,
possibly sanitizing them if they are present.

@@

The \code{binom} package is a popular R package for computing confidence
intervals, and the \code{binom.profile} function does so using the profile
likelihood.  This code snippet highlights a data sanitization pattern: the
programmer first binds the vectors into a matrix, with one column for each
of \code{x} and \code{n}, then finds rows where both columns are not \NA,
then extracts non-\NA values and stores them into \code{x} and \code{n}
respectively.

\paragraph{Scalars}
The data also suggests that programmers often use scalars, and do
dimensionality checks on their data.

@@

The \code{hankel.matrix} function takes two arguments.  The function has a
number of interesting checks: \code{n} undergoes an integer type check;
\code{x} undergoes a vector type check; and \code{n} is expected to be a
scalar, as if it were a vector, the code \code{m < n} would essentially
check \code{m < n[i]} for each element in \code{n}, returning a vector of
comparison results, and the conditional of an if must be a scalar else R
generates a warning.

\paragraph{Matrices}

Matrices represent another layer of complexity on top of vectorized
primitives.  Internally, matrices are simply vectors with class ``matrix''
and a \code{dims} attribute indicating matrix dimensions, and while not
codified in the language semantics, many internal functions will coerce
vectors to matrices automatically.  Relying on automatic coercion is unwise,
and many functions check to ensure that their arguments are matrices.

@@

@@

Thus far we've seen instances where we introduce types that subsume existing
argument validation code, but our intention is not to subsume {\it all} such
code.

@@

@@

\paragraph{Structs.}

We observed the use of consistently-named lists in \AT{a number of
  functions}.  In R, ``fields'' of named lists can be accessed using the
\code{$} operator.

@@

@@

We still believe struct types to be useful to R programmers, and so our
language supports the type and \contractr will check them, though we
configured \typetracer to ignore them as they made the data substantially
less useful.  On the positive side, we found that \AT{get the number, but
  often} structs-typed values had a {\it class}, which helped motivate our
inclusion of class-like types in our type annotation language.

%
\paragraph{Data Frames}

One of the most popular classes in R is the \code{data.frame} class.  Data
frames and the derivative \code{tibble} (from the \code{tidyverse}
ecosystem) and \code{data.table} (from the \code{data.table} package) types
underpin a huge amount of data analysis in R.

One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise into \typetracer's analysis results.  Further
complicating data frames is that many functions built to operate on them
operate in a name-agnostic way.  For instance, the \code{tidyverse} package
ecosystem allows programmers to pass column names to functions which operate
on their data frames.  In base R, idiomatic data frame use is to use string
column names to select rows from the frame (unless only a single column is
of interest, wherein the \code{$} syntax is appropriate).

@@

@@

\paragraph{Objects}

@@

\subsection{Robustness}

In the last section, we evaluated our type annotation language based on a
set of function types that were automatically generated by \typetracer.
Now, we're interested in knowing how representative our trace-generated
types are of the functions they were generated for.  To measure this, we
conducted a large-scale experiment: for each package in the corpus discussed
in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
vignette code, obtaining types for package functions, and then subsequently
ran the test, example, and vignette code of each of the package's reverse
dependencies (or clients).  We collected failed contract assertions in order
to analyze them and perhaps identify weaknesses in our type annotation
language.

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
of contract assertions failed.  We made a simplification in our analysis,
capping the number of function arguments we generate types for to 20, and
this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
that the simplification we made did not greatly affect the results.  Once we
controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
failed for reasons related to our generated types.  Additionally, we found
that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
were never violated.  The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we count out S3 functions: These S3 functions
represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures
generated by running \typetracer on our corpus are indeed representative,
and capture intended function behaviour.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  By far the most commonly failed assertion
is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf
  class}\langle matrix \rangle$.  This was a deliberate design decision, as
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested
in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing
contract check.  Digging a little deeper, we noticed that many of these
functions were performing S3 dispatch, which is part of R's
object-orientation framework which we will tackle in future work.  Removing
those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions
exhibited failing contract checks.  These are functions which were
under-tested, as calls to these functions represent only
\PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the
core corpus to infer types.
